Monitoring and Projecting Global Hunger

### 2020-12-10 docs/modeling_fies.tex
update figures; incorporate Homi changes

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{The World Data Lab, Vienna, Austria}
\affil[3]{Department of Economics, Vienna University of Economics and Business}
\affil[4]{Statistics Division, Food and Agricultural Organization of the United Nations}
\affil[5]{International Institute for Applied Systems Analysis}
\affil[6]{Wittgenstein Centre for Demography and Global Human Capital}
\affil[7]{Austrian Institute of Economic Research}
\affil[8]{The Brookings Institute}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

@@

\section{Main}
Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of at least moderate food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population. These do not directly capture the actual lived exposure of individuals to food shortages.

@@

Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.  However, for both indicators, these changes are very slight: in 2030, the number of moderately food insecure people will have increased by 2.2\% since 2010, and the number of severely food insecure people will have decreased by 7.0\%.  The increases in moderate food insecurity have been in lower and lower-middle income countries, while the increases in severe food insecurity are entirely from lower income countries.  For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}), we find clear improvements in all except high income countries.  Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2030s.  Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in 2020s.  In spite of this progress in the bottom three income quartiles, in high-income countries, the rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

@@

For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig. \ref{fig:vimp}), followed by the rate of stunting and the Gini coefficient.  For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.  For the rate of severe food insecurity, the country's Gini coefficient was more important compared to moderate food insecurity.  Across both indicators, economic variables were the most relevant predictors.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of progress in developing countries.  Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of food insecure people.

@@

While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate food insecurity to increase and the number of people experiencing severe food insecurity to fall only slightly between 2010 and 2030.  This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, a falling far short of the SDG 2 goal of ending hunger.  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, sustainable diets, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done at the political level in pursuit of SDG 2.

@@

For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions up to variables for 2030. We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.  This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.

@@

After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We use a common method of permuting each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.  This metric can provide some interpretability to model predictions.

>>>>

\affil[1]{T.H. Chan School of Public Health, Harvard University}
\affil[2]{The World Data Lab, Vienna, Austria}
\affil[3]{Department of Economics, Vienna University of Economics and Business}
\affil[4]{Statistics Division, Food and Agricultural Organization of the United Nations}
\affil[5]{International Institute for Applied Systems Analysis}
\affil[6]{Wittgenstein Centre for Demography and Global Human Capital}
\affil[7]{Austrian Institute of Economic Research}
\affil[8]{The Brookings Institution}
\affil[*]{Corresponding Author: mcooper@hsph.harvard.edu}

@@

\section{Main}
Food security is a critical component of human flourishing and its importance as a global policy objective is reflected in the second Sustainable Development Goal (SDG 2), ``Zero Hunger." One of the indicators to track progress on the first target of this goal, to ensure access by everyone to safe, nutritious and sufficient food (SDG2.1), is the prevalence of moderate and severe food insecurity in the population based on the Food Insecurity Experience Scale, or FIES. The FIES was developed by FAO in 2013 as a global extension of pioneering work done on Latin America to develop a metric of food insecurity based on the perspective of people who experience hunger and conditions that produce hunger. Other indicators of food insecurity, such as FAO's most commonly used estimate of undernourishment, are based on macro models of the mean and distribution of calories available per capita in a population. These do not directly capture the actual lived exposure of individuals to food shortages.

@@

Under realistic scenario assumptions, we find divergent trends in the number of moderate and severely food insecure people around the world, with the number of people at least moderately food insecure increasing and the number of people severely food insecure declining.  However, for both indicators, these changes are very slight: in 2030, the number of moderately food insecure people will have increased by 2.2\% since 2010, and the number of severely food insecure people will have decreased by 7.0\%.  The increases in moderate food insecurity have been in lower and lower-middle income countries, while the decreases in severe food insecurity are entirely from lower income countries.  For both the moderate and severe food insecurity thresholds, COVID-19 has had a clearly discernible impact, particularly in middle-income countries.

@@

Examining the rate of food insecurity in a population, rather than the total number of food insecure people (See Fig. \ref{fig:rates}), we find clear improvements in all except high income countries.  Countries that are currently upper-middle income, like China and Brazil, saw large gains in improving food insecurity rates throughout the 2010s, and will continue to see somewhat diminished gains in improving food security throughout the 2020s.  Lower and lower-middle income countries, on the other hand, saw more modest gains in the 2010s but are forecasted to see greater gains in reducing the rate of food insecurity in the 2020s.  In spite of this progress in the bottom three income quartiles, in high-income countries, the rates of people in moderate and severe food insecurity have been flat and are expected even to increase slightly.  Overall, the world has made steady but slow progress on reducing the rate of food insecurity.

@@

For severe food insecurity, the most important variable for predicting food security levels is the Poverty Headcount Index (See Fig. \ref{fig:vimp}), followed by the rate of stunting and the Gini coefficient.  For predicting the rate of people with at least moderate food insecurity, GDP per capita is the most important, followed closely by the Poverty Headcount Index.  The country's Gini coefficient is also important, more so for predicting the rate of severe food insecurity compared to moderate food insecurity.  Across both outcomes, economic variables were the most relevant predictors.

@@

\section{Discussion}
This paper presents the first global estimates of the experience of food insecurity at a subnational level, with forecasts until the year 2030.  We find that, while food insecurity has been increasing and has likely worsened in 2020 due to the economic downturn associated with the COVID-19 pandemic, our models predict that the percent of people who experience food insecurity globally will decrease throughout the 2020s, largely because of economic growth in developing countries.  Nevertheless, lower and lower-middle income countries are expected to see overall increases in the number of moderately food insecure people.

@@

While we find that rates of food insecurity are expected to decrease overall by 2030, our model, which relies on middle-of-the-road assumptions, still expects the number of people experiencing moderate food insecurity to increase and the number of people experiencing severe food insecurity to fall only slightly between 2010 and 2030.  This still leaves billions of people eating less than they should and three-quarters of a billion people going entire days without eating a decade from now, falling far short of the SDG 2 goal of ending hunger.  Moreover, decreases in food insecurity as measured by the FIES do not necessarily mean there will be improvements in other significant challenges such as poor dietary quality, micronutrient deficiencies, sustainable diets, or obesity.  Thus, while expected trends in correlates of food security give cause for optimism, there is still significant work to be done at the political level in pursuit of SDG 2.

@@

For health variables that have historically shown long-term trends and represent a rate of occurrence in a population, including stunting, wasting, and malaria mortality rate, we estimate the Annualized Rate of Change (AROC) for each subnational area to model these and to obtain predictions for variables up to 2030. We calculate the rate of change between each pair of years in the dataset, and then use the mean rate of change over the period for which data are available, giving greater weight to more recent years, and applying that mean rate of change to estimate future levels.  This method is used to obtain forecasts of the rates of stunting and wasting up to the year 2030.

@@

After fitting the models, we examine the importance of the individual covariates in explaining rates of moderate and severe food insecurity. We use a common method of permutating each individual variable and re-running the regressions, and then determining the difference in MAE between the regression \citep{ishwaran2007variable, breiman2001random}.  This metric can provide some interpretability to model predictions.
